### 2009-06-16 paper.tex
paper updates

We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
Theory predicts that ML inference, using the correct model and parameters, should maximize the perfect alignment rate.
Inspecting \figref{perfect}, we find this to be the case, except where the outgroup is very distant and the bins are probably undersampled
(when the perfect alignment rate drops below $\sim 5/125$, 125 trials may be too few).
We also note that the ML version of \stemloc\ is near-optimal, and slightly better than the posterior-decoding \stemloc-AMA.
Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods,
with no decisive advantage shown by either TKF91 (linear gap penalty) or Long Indel (affine gap penalty).

A subtly different trend is apparent when we look at the accuracy of the returned alignment,
measured not by the all-or-nothing metric of whether the alignment is {\em perfect},
but rather by a metric that count the number of correct alignment columns.
\figref{ama} shows the {\em Alignment Metric Accuracy} (AMA),
a generalization of the {\em Sum-of Pairs Score} (SPS)
which measures the proportion of residues which are correctly aligned, averaged over all pairs of sequences \cite{SchwartzMyersPachter2006}.
Here, we see that \stemlocama---which attempts to find the alignment with maximum expected AMA---edges out both
\indiegram\ and the ML version of \stemloc.
The decision-theoretic scoring scheme of \stemlocama\ appears to offset the minor incorrectness of its model.

Again, when looking at the AMA statistics, the most important factor is clearly the incorporation of some form of basepair structure
into the model: \handel, whether using TKF91 or the Long Indel model, performs much worse than the SCFG-based methods.
This is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
correctly aligned for free.

>>>>

We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
Theory predicts that ML inference, using the correct model and parameters, should maximize the perfect alignment rate.
Inspecting \figref{perfect}, we find this to be the case, except where the outgroup is very distant and the bins are probably undersampled
(when the perfect alignment rate drops below $\sim 5/125$, 125 trials may be too few).
We also note that the ML version of \stemloc\ is near-optimal, despite the \stemloc\ pair-SCFG being
slightly different from the TKFST pair-SCFG in parameterization and structure
 (e.g. \stemloc's grammar does not allow insertion/deletion of entire substructures).
The ML version of \stemloc\ is also observed to have a slightly higher perfect alignment rate than the posterior-decoding version (\stemloc-AMA).
Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods,
with no decisive advantage shown by either TKF91 (linear gap penalty) or Long Indel (affine gap penalty).

A subtly different ranking emerges from consideration of the {\em alignment accuracy}.
In \figref{ama}, we abandon the all-or-nothing metric of counting only perfect alignments,
instead using a metric that shows what proportion of the alignment is correct.
Specifically, we plot the {\em Alignment Metric Accuracy} (AMA) as a function of outgroup branch length.
AMA measures the proportion of residues which are correctly aligned, averaged over all pairs of sequences \cite{SchwartzMyersPachter2006}.
\figref{ama} reveals that \stemlocama\ (which attempts to find the alignment with the maximum expected AMA)
edges out both \indiegram\ and the ML version of \stemloc\ (both of which attempt to find the alignment with the maximum likelihood).
These results, compared to the subtly different story told by the perfect alignment rate,
underscore the point that benchmark results for alignment methods can depend exquisitely on the choice of accuracy metric.
The superiority of ML methods is only assured in terms of perfect alignment rate, and not necessarily other accuracy metrics.

With the AMA statistic, as with the perfect alignment rate, the most important factor distinguishing the various models
is clearly the incorporation of some form of basepair structure:
structure-less \handel\  (regardless of linear {\em vs} affine gap penalty) performs much worse than the structure-aware SCFG methods.
Intuitively, this is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
correctly aligned for free.
